Compositional Certified Resource Bounds

### 2014-07-03 paper/main.tex
fixed some typos

Following the development steps of automatic amortized analysis for
functional programs~\cite{Jost03,HoffmannH10}, we deliberately
restrict ourselves to linear bounds in the automatic analysis (there
are no restrictions for the manually-derived bounds in the logic).
More specifically, bounds have the form $\sum_{a,b} q_{(a,b)} |[a,b]|$
where $a$ and $b$ are integer variables or constants.  The reason for
our focus on linear bounds is mainly clarity of presentation.  We
already experimented with an extension to multivariate resource
polynomials~\cite{HoffmannAH11} but this would make the inference
rules considerably more involved and should better be described
separately.  However, we developed the linear inference system so that
the extension to polynomial bounds shall work smoothly.

To the best of our knowledge, this article presents the first resource
analysis framework for C that makes it possible to combine non-trivial
automatically derived bounds with interactively derived bounds in a
proof system that produces verifiable certificates for the bounds.
Our approach complements existing work since it provides a semantic
foundation for the computation of bounds while still providing support
for automation that sometimes goes beyond the capabilities of existing
techniques.  Both the automatic amortized analysis and the
quantitative logic are naturally compositional and describe the
resource behavior of code fragments without referring to the source
code.  As a result, we naturally derive global bounds that are
functions of the input parameters of the program.  Another unique
feature of our system is, that it can naturally handle resources like
memory that may become available during the execution.

We have evaluated the automatic analysis with system code and examples
from the literature.  \pref{app:cat} contains more than $30$
challenging loop and recursion patterns that we collected from open
source software and the literature.  Our analysis can find
asymptotically tight bounds for $30$ all but 3 of these patterns and
in most cases the derived constant factors are tight.  To compare our
automatic analyzer with existing techniques, we tested our examples
with the tools KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and
LOOPUS~\cite{SinnZV14}.  We also contacted the authors of
SPEED~\cite{GulwaniMC09} but were able to obtain a tool to perform a
comparison.  Our experiments shows that the bounds that we derive are
often more precise than those derived by existing tools.  Only
LOOPUS~\cite{SinnZV14}, which also uses amortization techniques, is
able to achieve a similar precision.

>>>>

Following the development steps of automatic amortized analysis for
functional programs~\cite{Jost03,HoffmannH10}, we deliberately
restrict ourselves to linear bounds in the automatic analysis (there
are no restrictions for the manually-derived bounds in the logic).
More specifically, bounds have the form $\sum_{a,b} q_{(a,b)} |[a,b]|$
where $a$ and $b$ are integer variables or constants.  The reason for
our focus on linear bounds is mainly clarity of presentation.  We
already experimented with an extension to multivariate resource
polynomials~\cite{HoffmannAH11,HoffmannS13} but this would make the inference
rules considerably more involved and should better be described
separately.  However, we developed the linear inference system so that
the extension to polynomial bounds shall work smoothly.

To the best of our knowledge, this article presents the first resource
analysis framework for C that makes it possible to combine non-trivial
automatically derived bounds with interactively derived bounds in a
proof system that produces verifiable certificates for the bounds.
Our approach complements existing work since it provides a semantic
foundation for the computation of bounds while still providing support
for automation that sometimes goes beyond the capabilities of existing
techniques.  Both the automatic amortized analysis and the
quantitative logic are naturally compositional and describe the
resource behavior of code fragments without referring to the source
code.  As a result, we naturally derive global bounds that are
functions of the input parameters of the program.  Another unique
feature of our system is, that it can handle resources like
memory that may become available during the execution.

We have evaluated the automatic analysis with system code and examples
from the literature.  \pref{app:cat} contains more than $30$
challenging loop and recursion patterns that we collected from open
source software and the literature.  Our analysis can find
asymptotically tight bounds for $30$ all but 3 of these patterns and
in most cases the derived constant factors are tight.  To compare our
automatic analyzer with existing techniques, we tested our examples
with the tools KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and
LOOPUS~\cite{SinnZV14}.  We also contacted the authors of
SPEED~\cite{GulwaniMC09} but were able to obtain a tool to perform a
comparison.  Our experiments show that the bounds that we derive are
often more precise than those derived by existing tools.  Only
LOOPUS~\cite{SinnZV14}, which also uses amortization techniques, is
able to achieve a similar precision.
